
Many thanks are due for making this PhD dissertation possible. 
%
First, I would like to thank my advisor, Cl\'audio Silva, for giving me the opportunity to pursue this research as well as his guidance throughout the entire PhD program. This opportunity changed my life, and for that I am very grateful.
%
I would also like to thank Luis Gustavo Nonato. Gustavo has helped me since the time of my master program at the University of S\~ao Paulo. He motivated me to join a PhD program and all his advice helped me during the course of my graduate program.  
%
I wish to thank Mike Kirby for our inspiring conversations and his guidance as well; they often led me to think about interesting problems, not only in sciences but in other areas as well. 
%
Cl\'audio, Gustavo and Mike taught me how to conduct research and to be critical of my own work.  
%
My appreciation also extends to Chris Johnson and Juliana Freire for being members of my committee and provide valuable comments. 

I would also like to acknowledge my co-authors. A special thanks to Jo\~ao Comba, Val\'erio Pascucci, Julien Tierny,  Lis Cust\'odio, Sin\'esio Pesco, Daniel J\"onson, and  Carlos Scheiddeger. In particular, I thank Carlos, with whom I had the opportunity to work with in most of the research presented in this dissertation. Carlos' passion is contagious and often made me challenge myself to learn more.
%
I am also very grateful for the wonderful people at the Scientific Computing and Imaging Institute (SCI) team, including professors, students and staff,  for providing a great learning environment at the University of Utah. 
%
At SCI, I had the opportunity to meet many brilliant people, from all around the world, and I deeply cherish their friendship.
%
A special thanks to Lauro Lins, Brian Summa, Carlos Dietrich, Josh Levine, Matt Berger, Attila Gyulassy, Emanuele Santos, Blake Nelson, and Joel Daniels for many stimulating and insightful conversations during the past four years.
%
Lastly, I would like to thank my parents for encouraging me throughout the years; and my wife and friends for being loving and supportive of my work and professional goals.